The socialization of meritocracy and market justice preferences at school

### 2024-10-02 paper-blinded.tex
Update revised version MPI Submit

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

@@

\begin{abstract}
Previous research has shown that schools justify student performance
differences based on meritocratic ideals. However, the link between
school meritocracy and students' justification of inequalities has
received little attention. This paper examines how school and family
socialization influence students' beliefs about meritocracy and
market-based justifications for access to health, pensions, and
education in Chile. Using data from the First Study of Civic Education
in Chile, we employ a multilevel modeling strategy to analyze 5,047
students and parents nested within 231 schools. Our findings reveal that
a significant proportion of Chilean students agree with market justice
principles, at a higher rate than adults. Most students endorse
meritocratic views, particularly the reward of effort. Multilevel
estimations show strong associations between meritocratic beliefs and
market justice preferences: students who believe in the reward of effort
and talent are more likely to justify inequalities based on financial
capacity. At the school level, market justice preferences are higher in
high-status schools but lower in schools with better achievement.
Additionally, the conditional influence of meritocracy beliefs weakens
in schools with higher socioeconomic status and performance. These
results suggest that meritocratic beliefs and market justice preferences
begin early and are influenced by the school environment.
\end{abstract}
\vskip 3em

@@

\begin{ideuno}

Front Matter

\end{ideuno}
% -------------------
% END TITLE SECTION
% -------------------

\begin{idedos}

Manuscript
\end{idedos}

\setstretch{1.15}
\section{Introduction}\label{introduction}

@@

In the present paper, we address the role of the perception of
meritocracy on the justification of social inequalities by eighth-grade
students in Chile, a country characterized by a highly stratified
educational system. In particular, we focus on justifying inequalities
in health, education and pensions, which are traditional social policy
areas where access to better services could be justified based on
payment capacity, referred to as market justice
{[}\citeproc{ref-lane_market_1986}{9},\citeproc{ref-lindh_bringing_2023}{10}{]}.
Most of the research in this realm has only considered adults, leaving
aside the study of how beliefs in this field develop at student age as
well as the impact of the school context and the family as the main
socialization agencies {[}\citeproc{ref-batruch_belief_2022}{3}{]}.
Regarding schools, the way in which they deal with unequal conditions of
origin has been linked to the \emph{hidden curriculum}
{[}\citeproc{ref-chafel_schooling_1997}{11}{]}, whereby students learn
about distributive norms in society and mechanisms of justification of
social differences.

@@

The Chilean case is particularly intriguing for studying market justice
preferences. This country is characterized by acute and persistent
economic inequality, which stands out in Latin America and among OECD
countries {[}\citeproc{ref-flores_top_2020}{12}{]}. In Chile, the
poorest 50\% captures only 10\% of the total income and has negative
wealth, while the richest 1\% receives almost 27\% of the income and
holds 49.6\% of the wealth {[}\citeproc{ref-chancel_world_2022}{13}{]}.
Much of this inequality has been attributed to the deep neoliberal
reforms that institutionalized privatization and commodification of
various economic sectors. Such reforms where introduced during the
dictatorship (1973-1989) and expanded in democracy through concessions,
demand credits, and specific regulatory frameworks
{[}\citeproc{ref-ffrench-davis_reformas_2018}{14}{]}. This shift in
economic policy has allowed the unprecedented emergence of markets in
areas related to social reproduction, such as health, pensions, and
education, with provision and access managed by private entities and
segmented by individual payment capacity, heavily reliant on State
subsidies {[}\citeproc{ref-boccardo_30_2020}{15}{]}. In health, although
the majority of the population uses the public insurance system
(78.9\%), 15.3\% are served by private insurers
{[}\citeproc{ref-observatoriosocial_estadisticas_2024}{16}{]}. The
pension system is based on individual capitalization, with mandatory
contributions managed by private administrators investing in the
financial market, currently involving 11 million contributors
{[}\citeproc{ref-superintendenciadepensiones_estadisticas_2024}{17}{]}.
In education, 30.6\% of school enrollment is in public schools, 54.0\%
in state-subsidized (voucher) private schools, and 9.3\% in fully
private schools, generally attracting higher-income groups
{[}\citeproc{ref-ministeriodeeducacion_resumen_2023}{18}{]}.

@@

The justification of social inequality based on market-type criteria has
been conceptualized as the individuals' adherence to the deservingness
of social goods and services (such as health, education, and pensions)
based on prices and individuals' ability to pay
{[}\citeproc{ref-lane_market_1986}{9},\citeproc{ref-boltanski_new_2005}{19},\citeproc{ref-streeck_citizens_2012}{20}{]}.
Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
{[}\citeproc{ref-kluegel_beliefs_1987}{21}{]}, highlights that people's
explanations and justifications of social inequality are closely tied to
their judgments of deservingness. The influence of ideologies
{[}\citeproc{ref-wegener_dominant_1995}{22}{]} and cultural schemas
{[}\citeproc{ref-homan_being_2017}{23}{]} is pivotal in shaping these
explanations by offering symbolic representations that frame societal
structures and expectations. While significant attention has been paid
to wage inequality, income distribution, and payment differentials in
the literature
{[}\citeproc{ref-castillo_legitimacy_2011}{24}--\citeproc{ref-shariff_income_2016}{27}{]},
there has been less examination of public beliefs about which life
domains should be governed by market relations
{[}\citeproc{ref-lindh_bringing_2023}{10}{]} and even less about
children's acceptance or rejection of these market principles. This
oversight is notable given the extensive encroachment of market logic
into public goods, welfare policy, and social services over the past
five decades
{[}\citeproc{ref-centeno_arc_2012}{28},\citeproc{ref-harvey_breve_2015}{29}{]},
affecting areas such as pensions, health services, and education.

There are substantial differences in funding and delivery methods in the
management of social services across nations
{[}\citeproc{ref-jensen_worlds_2008}{30},\citeproc{ref-stoy_worlds_2014}{31}{]}.
Nordic countries, for example, predominantly employ public agencies to
produce and provide social services, funding these through collective
taxation and offering them in kind to the majority of citizens. This
system prioritizes social justice, placing it above market mechanisms in
accessing services. In contrast, other countries rely more heavily on
for-profit entities and private funding, where service distribution
depends mainly on individual financial capacity to pay user fees,
highlighting the influence of market justice in service allocation. The
trend toward marketization of welfare services has been growing since
the 1980s {[}\citeproc{ref-salamon_marketization_1993}{32}{]}, and this
shift is increasingly evident even in countries where market solutions
have traditionally had a minor role in social policy
{[}\citeproc{ref-sivesind_changing_2017}{33}{]}. The expansion of
marketization has been related to a larger justification of market
mechanisms, whereby societies with larger private spending on services
show larger market justice preferences
{[}\citeproc{ref-lindh_public_2015}{34}{]}.

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
{[}\citeproc{ref-lane_market_1986}{9}{]}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
{[}\citeproc{ref-greitemeyer_subjective_2016}{35}--\citeproc{ref-power_deprivationprotest_2018}{38}{]},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see {[}\citeproc{ref-wilson_role_2003}{39}{]}).

Despite high-income inequality and limited social mobility in Chile, and
in Latin America in general, there is a prevalent belief that
individuals are solely responsible for their economic outcomes, a view
that varies across the region
{[}\citeproc{ref-bucca_merit_2016}{40}--\citeproc{ref-salgado_inequality_2023}{43}{]}.
The reliance on private welfare providers and widespread user fees
{[}\citeproc{ref-molyneux_neoliberal_2008}{44}{]} adds complexity to
this context, as reflected in surveys conducted by the Center for Public
Studies (CEP). According to this data, 35.9\% prefer private health
insurance, and 63\% would prefer private education
{[}\citeproc{ref-centrodeestudiospublicos_estudio_2024}{45}{]}. Yet,
research on children's justification in this area remains limited,
highlighting a significant gap in understanding how younger generations
view market-based access to welfare and whether these views are
associated with their meritocratic beliefs.

@@

The original definition of merit is a combination of effort and talent
{[}\citeproc{ref-young_rise_1958}{5}{]}, and a meritocracy is a
distributive system where merit is the main criterion for allocating
valuable goods and rewards. From a sociological perspective, meritocracy
has been used in research on social mobility to characterize societies
with low mobility that threaten the meritocratic ideal
{[}\citeproc{ref-goldthorpe_myth_2003}{46}{]}. More recently, sociology
and social psychology research has attended to subjective aspects
related to the support for meritocratic principles in different
societies, such as beliefs in meritocracy
{[}\citeproc{ref-castillo_multidimensional_2023}{47}--\citeproc{ref-mijs_paradox_2019}{49}{]}.

Meritocratic beliefs can cover two types of subjective processes:
preferences and perceptions
{[}\citeproc{ref-castillo_multidimensional_2023}{47}{]}. While
meritocratic preferences refer to a justification of distribution based
on merit criteria (effort and talent), the perception of meritocracy
refers to how individuals view and understand the concept of meritocracy
in their society
{[}\citeproc{ref-duru-bellat_who_2012}{50},\citeproc{ref-castillo_meritocracia_2019}{51}{]}.
The perception can vary greatly depending on individual experiences, as
well as social, economic, and cultural background. Some people may see
meritocracy as a fair and just system that allows anyone to succeed
based on their abilities and hard work. In contrast, others may view it
as a myth or a cover for existing power dynamics and inequality, serving
to maintain and even reinforce inequality
{[}\citeproc{ref-mijs_paradox_2019}{49},\citeproc{ref-lampert_meritocratic_2013}{52}{]}.
Some studies have analyzed how those with greater privileges believe
more in meritocracy {[}\citeproc{ref-reynolds_perceptions_2014}{53}{]},
how greater economic inequality increases meritocratic beliefs
{[}\citeproc{ref-mijs_paradox_2019}{49}{]}, and how larger inequality
decreases it {[}\citeproc{ref-morris_representing_2022}{54}{]}.

@@

\(H_{1a}\): Students who perceive that there is more meritocracy in
society will show larger market justice preferences

@@

\(H_{1b}\): Students who perceive that there is more meritocracy at
school will show larger market justice preferences

@@

\(H_2\): Students from families of higher social status will show larger
market justice preferences.

@@

\(H_3\): Students from schools of higher social status will show larger
market justice preferences.

\(H_4\): Students from schools with an average better performance in
standardized achievement tests will show larger market justice
preferences.

@@

\(H_5\): The family's social status will moderate the effect of
meritocratic perception on market justice preferences.

\(H_6\): The school's status will moderate the effect of meritocratic
perceptions on market justice preferences.

\(H_7\): The school's performance in standardized achievement tests will
moderate the effect of meritocratic perceptions on market justice
preferences.

@@

\section{Data, variables and method}\label{data-variables-and-method}

@@

The First Study of Civic Education in Chile, conducted by the Agency for
Quality Education of the Ministry of Education in 2017, serves as the
main data source. This database is composed by a civic knowledge test
score and a series of items batteries measuring different aspects of
citizenship. The target population includes 8th-grade students in 242
schools nationwide, as well as their parents and teachers. Initially,
the database contains 8,589 student observations and 6,770 parent
observations. To ensure higher data quality and considering the survey's
unit of analysis, we removed 171 student cases and 79 parent cases that
exhibited repetitive and careless response patterns
{[}\citeproc{ref-gottfried_autocorrelation_2022}{73}{]}. Additionally,
we utilized data from the System for Measuring the Quality of Education
(SIMCE) conducted by the Ministry of Education in 2017, which provides
school-level information such as administrative dependence,
socioeconomic classification, and results obtained in standardized
mathematics and language tests. After processing the variables and
removing missing cases, the final database for this study includes a
two-level stratified sample composed of 5,047 students and parents
(level 1) nested within 231 schools (level 2).

@@

\textbf{Market Justice Preferences}: The dependent variables in this
study are market justice preferences. This construct is measured through
three variables that address the degree of justification regarding
whether access to social services in pensions, education, and health
should be income conditional. Justification of inequality in health is
measured by the item: ``Is it fair in Chile that people with higher
incomes can access better healthcare than those with lower incomes?''
The same question is asked for education and pensions. In all cases,
respondents indicate their preferences on a Likert scale ranging from
``completely disagree'' (1) to ``completely agree'' (4). Additionally,
we include a summarized indicator of ``market justice preferences'',
measured by an average index across these items (α = 0.86), with values
ranging from 1 to 4, where higher values represent greater preferences
for market justice (see Table~\ref{tbl-desc-dependientes}). We analyzed
these items independently as well as by the average index.

@@

This study focuses on two school-level characteristics: socioeconomic
status and academic performance. Socioeconomic status is assessed using
the Ministry of Education's classification, measured as an ordinal item
with five categories ranging from ``low'' (1) to ``high'' (5). Academic
performance is measured using the school's results in the SIMCE (System
of Measurement of Educational Quality) standardized tests, which are
administered yearly at different educational levels in areas such as
language and mathematics. These results are categorized as ``low,''
``medium,'' and ``high.'' The contextual level items used, response
categories, and their frequencies are detailed in
Table~\ref{tbl-desc-school}.

@@

The hypotheses of this research were pre-registered in the Open Science
Framework platform of the Center for Open Science (OSF), the access to
the document is available at this
\href{https://doi.org/10.17605/OSF.IO/UFSDV}{link}. The statistical
analysis for this research was conducted using the \emph{lme4} package
in R version 4.1.3.

@@

\subsection{Descriptive}\label{descriptive}

@@

\subsection{Multivariate}\label{multivariate}

@@

Model 1 in Table~\ref{tbl-ordinal-reg} for the three dependent variables
of justice in differential access to health, pensions, and education
demonstrates a similar trend regarding the effect of the perception of
meritocracy on these items, with all effects being statistically
significant (\emph{p} \textless{} 0.05). Regarding the variables of
social meritocracy, the three variables---talent, effort, and
deservingness---show that as these increase, the justification for
differentiated access to pensions, education, and health also increases.
In the context of school meritocracy, the effects are mixed: as
perceived school talent increases, the justification for differentiated
access to these services increases; conversely, as perceived school
effort increases, the justification decreases, holding all other
variables constant.

Additionally, Model 1 examines family status through parents'
educational level and the number of books at home. For health, pensions,
and education, a negative but statistically non-significant association
with parents' education is observed. However, having more than 25 books
at home shows a negative and statistically significant effect on market
justice in health and pensions, suggesting less justification for
differentiated access compared to those with fewer books.

Model 2 in Table~\ref{tbl-ordinal-reg} incorporates school-level
variables. The results show that high socioeconomic level schools show
greater preferences for market justice in health, pensions, and
education compared to low socioeconomic level schools, with
statistically significant results (\emph{p} \textless{} 0.05). Moreover,
and contrary to our hypothesis, schools with higher academic performance
show negative and statistically significant effects across the three
items.

@@

Table~\ref{tbl-interact} displays the interaction terms related to our
hypotheses. The results suggest that students' family background
moderates the relationship between their meritocratic perceptions and
their justification of market allocation in social services.
Specifically, the relationship between social effort and market justice
preferences is less positive for students whose parents have university
or postgraduate education. Lower-status students (measured by parental
education) prefer more market justice when they strongly adhere to
meritocratic perceptions of deservingness. A similar moderating effect
is seen with family cultural capital: the relationship between effort
and market justice preferences is less positive for students with more
than 25 books at home.

@@

\begin{idetres}

Back Matter
\end{idetres}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\begin{abstract}
Previous research has shown that schools often justify student
performance differences using meritocratic ideals. One potential
consequence of such ideals is the legitimization of outcome inequalities
across various spheres, including those traditionally associated with
equality and redistribution. In this study, we argue that the promotion
of meritocratic values during school-age can shape students' beliefs
about meritocracy and influence their views on market-based access to
health, pensions, and education. Using data from the 2017 National Study
of Civic Education in Chile, which includes 5,047 8th-grade students
from 231 schools, we estimated a series of multilevel models (lme4
library, R version 4.1.3) to test our hypotheses. Our findings show that
a significant proportion of Chilean students agree with market justice
principles---more so than adults. Most students endorse meritocratic
views, particularly the notion that effort should be rewarded, which
strongly correlates with market justice preferences: students who
believe in meritocracy are more likely to justify inequalities based on
financial capacity. At the school level, market justice preferences are
higher in high-status schools but lower in schools with higher academic
achievement. Furthermore, the conditional influence of meritocratic
beliefs diminishes in schools with higher socioeconomic status and
performance levels. These results suggest that the association between
meritocratic beliefs and market justice preferences is already
established at school age and is shaped by the school environment.
\end{abstract}
\vskip 3em

@@

% -------------------
% END TITLE SECTION
% -------------------

\setstretch{1.15}
\newpage{}

\section{Introduction}\label{introduction}

@@

In the present paper, we address the role of the perception of
meritocracy on the justification of social inequalities by eighth-grade
students in Chile, a country characterized by a highly stratified
educational system. In particular, we focus on justifying inequalities
in health, education and pensions, which are traditional social policy
areas where access to better services could be justified based on
payment capacity, referred to as market justice
{[}\citeproc{ref-lane_market_1986}{9},\citeproc{ref-lindh_bringing_2023}{10}{]}.
Market justice preferences refers to a belief in a market-based
distributive principle, where goods and rewards are allocated according
to individual merit, effort and contributions in a free-market economy
{[}\citeproc{ref-lindh_public_2015}{11}{]}.

Most of the research on meritocracy and justice preferences has only
considered adults, leaving aside the study of how beliefs in this field
develop at student age as well as the impact of the school context and
the family as the main socialization agencies
{[}\citeproc{ref-batruch_belief_2022}{3}{]}. Studying beliefs about
meritocracy and market justice preferences at school age is crucial
because these early beliefs can shape individuals' attitudes towards
inequality and social justice later in life. As Castillo et al.
{[}\citeproc{ref-castillo_meritocracia_2019}{12}{]} argues, meritocratic
ideals serve as a principle that legitimizes the distribution of goods
and rewards based on individual talent and effort, making it essential
to understand how these beliefs develop during adolescence. Furthermore,
early exposure to these beliefs may influence broader perceptions of
fairness, as Reynolds and Xian
{[}\citeproc{ref-reynolds_perceptions_2014}{13}{]} highlights, beliefs
about the stratification system influence people's judgments about the
fairness of inequality, suggesting that such attitudes formed at a young
age can have lasting impacts. Additionally, García-Sierra
{[}\citeproc{ref-garcia-sierra_dark_2023}{14}{]} emphasizes that the
effects of holding meritocratic beliefs are not uniform, as these
ideologies may reinforce socioeconomic disparities depending on
students' social backgrounds.

The Chilean case is particularly intriguing for studying market justice
preferences. This country is characterized by acute and persistent
economic inequality, which stands out in Latin America and among OECD
countries {[}\citeproc{ref-flores_top_2020}{15}{]}. In Chile, the
poorest 50\% captures only 10\% of the total income and has negative
wealth, while the richest 1\% receives almost 27\% of the income and
holds 49.6\% of the wealth {[}\citeproc{ref-chancel_world_2022}{16}{]}.
Much of this inequality has been attributed to the deep neoliberal
reforms that institutionalized privatization and commodification of
various economic sectors. Such reforms where introduced during the
dictatorship (1973-1989) and expanded in democracy through concessions,
demand credits, and specific regulatory frameworks
{[}\citeproc{ref-ffrench-davis_reformas_2018}{17}{]}. This shift in
economic policy has allowed the unprecedented emergence of markets in
areas related to social reproduction, such as health, pensions, and
education, with provision and access managed by private entities and
segmented by individual payment capacity, heavily reliant on State
subsidies {[}\citeproc{ref-boccardo_30_2020}{18}{]}. In health, although
the majority of the population uses the public insurance system
(78.9\%), 15.3\% are served by private insurers
{[}\citeproc{ref-observatoriosocial_estadisticas_2024}{19}{]}. The
pension system is based on individual capitalization, with mandatory
contributions managed by private administrators investing in the
financial market, currently involving 11 million contributors
{[}\citeproc{ref-superintendenciadepensiones_estadisticas_2024}{20}{]}.
In education, 30.6\% of school enrollment is in public schools, 54.0\%
in state-subsidized (voucher) private schools, and 9.3\% in fully
private schools, generally attracting higher-income groups
{[}\citeproc{ref-ministeriodeeducacion_resumen_2023}{21}{]}.

@@

@@

The justification of social inequality based on market-type criteria has
been conceptualized as the individuals' adherence to the deservingness
of social goods and services (such as health, education, and pensions)
based on prices and individuals' ability to pay
{[}\citeproc{ref-lane_market_1986}{9},\citeproc{ref-boltanski_new_2005}{22},\citeproc{ref-streeck_citizens_2012}{23}{]}.
Research on social stratification beliefs, which explore individual
perceptions of who deserves what and why
{[}\citeproc{ref-kluegel_beliefs_1987}{24}{]}, highlights that people's
explanations and justifications of social inequality are closely tied to
their judgments of deservingness. The influence of ideologies
{[}\citeproc{ref-wegener_dominant_1995}{25}{]} and cultural schemas
{[}\citeproc{ref-homan_being_2017}{26}{]} is pivotal in shaping these
explanations by offering symbolic representations that frame societal
structures and expectations. While significant attention has been paid
to wage inequality, income distribution, and payment differentials in
the literature
{[}\citeproc{ref-castillo_legitimacy_2011}{27}--\citeproc{ref-shariff_income_2016}{30}{]},
there has been less examination of public beliefs about which life
domains should be governed by market relations
{[}\citeproc{ref-lindh_bringing_2023}{10}{]} and even less about
children's acceptance or rejection of these market principles. This
oversight is notable given the extensive encroachment of market logic
into public goods, welfare policy, and social services over the past
five decades
{[}\citeproc{ref-centeno_arc_2012}{31},\citeproc{ref-harvey_breve_2015}{32}{]},
affecting areas such as pensions, health services, and education.

There are substantial differences in funding and delivery methods in the
management of social services across nations
{[}\citeproc{ref-jensen_worlds_2008}{33},\citeproc{ref-stoy_worlds_2014}{34}{]}.
Nordic countries, for example, predominantly employ public agencies to
produce and provide social services, funding these through collective
taxation and offering them in kind to the majority of citizens. This
system prioritizes social justice, placing it above market mechanisms in
accessing services. In contrast, other countries rely more heavily on
for-profit entities and private funding, where service distribution
depends mainly on individual financial capacity to pay user fees,
highlighting the influence of market justice in service allocation. The
trend toward marketization of welfare services has been growing since
the 1980s {[}\citeproc{ref-salamon_marketization_1993}{35}{]}, and this
shift is increasingly evident even in countries where market solutions
have traditionally had a minor role in social policy
{[}\citeproc{ref-sivesind_changing_2017}{36}{]}. The expansion of
marketization has been related to a larger justification of market
mechanisms, whereby societies with larger private spending on services
show larger market justice preferences
{[}\citeproc{ref-lindh_public_2015}{11}{]}.

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
{[}\citeproc{ref-lane_market_1986}{9}{]}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
{[}\citeproc{ref-greitemeyer_subjective_2016}{37}--\citeproc{ref-power_deprivationprotest_2018}{40}{]},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see {[}\citeproc{ref-wilson_role_2003}{41}{]}).

Despite high-income inequality and limited social mobility in Chile, and
in Latin America in general, there is a prevalent belief that
individuals are solely responsible for their economic outcomes, a view
that varies across the region
{[}\citeproc{ref-bucca_merit_2016}{42}--\citeproc{ref-salgado_inequality_2023}{45}{]}.
The reliance on private welfare providers and widespread user fees
{[}\citeproc{ref-molyneux_neoliberal_2008}{46}{]} adds complexity to
this context, as reflected in surveys conducted by the Center for Public
Studies (CEP). According to this data, 35.9\% prefer private health
insurance, and 63\% would prefer private education
{[}\citeproc{ref-centrodeestudiospublicos_estudio_2024}{47}{]}. Yet,
research on children's justification in this area remains limited,
highlighting a significant gap in understanding how younger generations
view market-based access to welfare and whether these views are
associated with their meritocratic beliefs.

@@

The original definition of merit is a combination of effort and talent
{[}\citeproc{ref-young_rise_1958}{5}{]}, and a meritocracy is a
distributive system where merit is the main criterion for allocating
valuable goods and rewards. From a sociological perspective, meritocracy
has been used in research on social mobility to characterize societies
with low mobility that threaten the meritocratic ideal
{[}\citeproc{ref-goldthorpe_myth_2003}{48}{]}. More recently, sociology
and social psychology research has attended to subjective aspects
related to the support for meritocratic principles in different
societies, such as beliefs in meritocracy
{[}\citeproc{ref-castillo_multidimensional_2023}{49}--\citeproc{ref-mijs_paradox_2019}{51}{]}.

Meritocratic beliefs can cover two types of subjective processes:
preferences and perceptions
{[}\citeproc{ref-castillo_multidimensional_2023}{49}{]}. While
meritocratic preferences refer to a justification of distribution based
on merit criteria (effort and talent), the perception of meritocracy
refers to how individuals view and understand the concept of meritocracy
in their society
{[}\citeproc{ref-castillo_meritocracia_2019}{12},\citeproc{ref-duru-bellat_who_2012}{52}{]}.
The perception can vary greatly depending on individual experiences, as
well as social, economic, and cultural background. Some people may see
meritocracy as a fair and just system that allows anyone to succeed
based on their abilities and hard work. In contrast, others may view it
as a myth or a cover for existing power dynamics and inequality, serving
to maintain and even reinforce inequality
{[}\citeproc{ref-mijs_paradox_2019}{51},\citeproc{ref-lampert_meritocratic_2013}{53}{]}.
Some studies have analyzed how those with greater privileges believe
more in meritocracy {[}\citeproc{ref-reynolds_perceptions_2014}{13}{]},
how greater economic inequality increases meritocratic beliefs
{[}\citeproc{ref-mijs_paradox_2019}{51}{]}, and how larger inequality
decreases it {[}\citeproc{ref-morris_representing_2022}{54}{]}.

@@

\(H_{1a}\): Students who perceive greater meritocracy in society will
exhibit stronger preferences for market justice.

@@

\(H_{1b}\): Students who perceive greater meritocracy in their schools
will exhibit stronger preferences for market justice.

@@

\(H_2\): Students from higher social status families will exhibit
stronger preferences for market justice.

@@

\(H_3\): Students from higher social status schools will show stronger
preferences for market justice.

\(H_4\): Students from schools with higher average performance on
standardized achievement tests will show stronger preferences for market
justice.

@@

\(H_5\): The family's social status will moderate the relationship
between perceptions of meritocracy and preferences for market justice.

\(H_6\): The school's status will moderate the relationship between
perceptions of meritocracy and preferences for market justice.

\(H_7\): The school's performance on standardized achievement tests will
moderate the relationship between perceptions of meritocracy and
preferences for market justice.

@@

\section{Data, Variables and Methods}\label{data-variables-and-methods}

@@

The First Study of Civic Education in Chile, conducted by the Agency for
Quality Education of the Ministry of Education in 2017, is the primary
data source. This database comprises a civic knowledge test score and a
series of items that measure different aspects of citizenship. The
target population includes 8th-grade students in 242 schools nationwide.
Initially, the database contains 8,589 student observations.To ensure
higher data quality and considering the survey's unit of analysis, we
removed 171 student cases that exhibited repetitive and careless
response patterns
{[}\citeproc{ref-gottfried_autocorrelation_2022}{73}{]}. Additionally,
we utilized data from the System for Measuring the Quality of Education
(SIMCE) conducted by the Ministry of Education in 2017, which provides
school-level information such as administrative dependence,
socioeconomic classification, and results obtained in standardized
mathematics and language tests. After processing the variables and
removing missing cases, the final database for this study includes a
two-level stratified sample composed of 5,047 students (level 1) nested
within 231 schools (level 2).

@@

\textbf{Market Justice Preferences}: The dependent variables in this
study are market justice preferences. This construct is measured through
three variables that address the degree of justification regarding
whether access to social services in pensions, education, and health
should be conditional on income. The Justification of inequality in
health is measured by the following item: ``Is it fair in Chile that
people with higher incomes can access better healthcare than those with
lower incomes?'' The same question is asked for education and pensions.
In all cases, respondents indicate their preferences on a Likert scale
ranging from ``completely disagree'' (1) to ``completely agree'' (4).
Additionally, we include a summarized indicator of ``market justice
preferences'', measured by an average index across these items (α =
0.86), with values ranging from 1 to 4, where higher values represent
stronger preferences for market justice (see
Table~\ref{tbl-desc-dependientes}). We analyzed these items
independently and by the average index.

@@

This study focuses on two school-level characteristics: socioeconomic
status and academic performance. Socioeconomic status is assessed using
the Ministry of Education's classification, measured as an ordinal item
with five categories ranging from ``low'' (1) to ``high'' (5). Academic
performance is measured using the school's results in the SIMCE (System
of Measurement of Educational Quality) standardized tests, administered
yearly at different educational levels in language and mathematics.
These results are categorized as ``low,'' ``medium,'' and ``high.'' The
contextual level items used, response categories, and their frequencies
are detailed in Table~\ref{tbl-desc-school}.

@@

The hypotheses of this research were pre-registered in the Open Science
Framework platform of the Center for Open Science (OSF), and access to
the document is available at this
\href{https://doi.org/10.17605/OSF.IO/UFSDV}{link}. The statistical
analysis for this research was conducted using the \emph{lme4} package
in R version 4.1.3.

@@

\subsection{Descriptive statistics}\label{descriptive-statistics}

@@

\subsection{Multivariate analysis}\label{multivariate-analysis}

@@

Model 1 in Table~\ref{tbl-ordinal-reg} for the three dependent variables
of justice in differential access to health, pensions, and education
demonstrates a similar trend regarding the effect of the perception of
meritocracy on these items, with all effects being statistically
significant (\emph{p} \textless{} 0.05). Regarding the variables of
social meritocracy, the three variables---talent (\(\beta\) = 0.16),
effort (\(\beta\) = 0.11), and deservingness (\(\beta\) = 0.51)---show
that as these increase, the justification for differentiated access to
pensions, education, and health also increases. In the context of school
meritocracy, the effects are mixed: as perceived school talent
increases, the justification for differentiated access to these services
increases (\(\beta\) = 0.23); conversely, as perceived school effort
increases, the justification decreases (\(\beta\) = -0.26), holding all
other variables constant.

Additionally, Model 1 examines family status through parents'
educational level and the number of books at home. For health, pensions,
and education, a negative but statistically non-significant association
with parents' education is observed. However, having more than 25 books
at home shows a negative and statistically significant effect on market
justice in health and pensions (\(\beta\) = -0.13; \emph{p} \textless{}
0.001), suggesting less justification for differentiated access compared
to those with fewer books.

Model 2 in Table~\ref{tbl-ordinal-reg} incorporates school-level
variables. The results show that high socioeconomic level schools show
greater preferences for market justice in health, pensions, and
education compared to low socioeconomic level schools, with
statistically significant results (\(\beta\) = 0.71; \emph{p}
\textless{} 0.05). Moreover, and contrary to our hypothesis, schools
with higher academic performance show negative and statistically
significant effects across the three items (\(\beta\) = -0.56; \emph{p}
\textless{} 0.001).

@@

Table~\ref{tbl-interact} displays the interaction terms related to our
hypotheses. The results suggest that students' family background
moderates the relationship between their meritocratic perceptions and
their justification of market allocation in social services.
Specifically, the relationship between social effort and market justice
preferences is less positive for students whose parents have university
or postgraduate education (\(\beta\) = -0.079; \emph{p} \textless{}
0.05). Lower-status students (measured by parental education) prefer
more market justice when they strongly adhere to meritocratic
perceptions of deservingness (\(\beta\) = -0.118; \emph{p} \textless{}
0.01). A similar moderating effect is seen with family cultural capital:
the relationship between effort and market justice preferences is less
positive for students with more than 25 books at home (\(\beta\) =
-0.081; \emph{p} \textless{} 0.01).

@@
